Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-22 paper.qmd
Update paper.qmd

---
title: "Nix for Polyglot, Reproducible Data Science Workflows"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
abstract: |
  Reproducible analysis requires more than clean, well-documented code: 
  researchers must also manage software dependencies, computational environments, and workflow execution. 
  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge, but no single solution handles polyglot environments, 
  system dependencies, and pipeline orchestration together. I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot 
  environments and declarative workflow orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define reproducible environments 
  spanning [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, and system dependencies. Building on these environments, [rixpress]{.pkg} orchestrates polyglot pipelines where each 
  computational step runs in its own hermetically sealed environment, with automatic caching and dependency tracking. 
  This approach enables researchers to develop analyses interactively while maintaining bit-for-bit reproducibility, supports 
  collaboration across heterogeneous computational environments, and ensures that analyses remain executable years into the future.

@@

However, [renv]{.pkg} has important limitations. It does not restore the version
of [R]{.proglang} itself—installing the correct R version must be done
separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
[renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg} cannot
resolve this conflict. Users must manually install system libraries, and the
process differs across operating systems. Further details are discussed in the
[renv]{.pkg} documentation^[(https://rstudio.github.io/renv/articles/renv.html#caveats)].

@@

For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, I refer readers to @dolstra2004nix.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, I recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nixnstaller}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`. This shell contains all the packages specified in `default.nix` and
can be used for development, similar to activating a virtual environment in the
[Python]{.proglang} ecosystem.

@@

The `rstats-on-nix` fork allows us to circumvent these limitations. For example, it provides
daily snapshots of CRAN. Each day, the [R]{.proglang} package set is updated and
committed to a dated branch using GitHub Actions. Users can select a specific
date with:

@@

I strive to provide an available date per week: each Monday, a GitHub Action
tests popular packages on Linux and macOS, and only if all tests succeed is the
date added to the list of available dates in [rix]{.pkg}. This ensures users can
reliably install packages, and allows us to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. I
backported the fix to the `rstats-on-nix` fork, making RStudio available to users of 
[rix]{.pkg} earlier than upstream, as merging PRs in the official repository can take some
time.

@@

To illustrate these capabilities with a realistic research task, I present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang} [@bezanson2017julia], uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report. The code to this example can be found in the  
[following repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
with additional details provided in the Appendix of this manuscript.

@@

Furthermore, [rixpress]{.pkg} extends this reproducibility to entire analysis
pipelines. By leveraging Nix as a build automation engine, [rixpress]{.pkg}
allows polyglot workflows where each step runs in its own hermetically sealed
environment. This ensures deep reproducibility, efficient caching, and seamless
orchestration of polyglot analyses, making it easier to manage complex
projects with confidence that results can be reproduced exactly.

@@

The source code of this paper is hosted on GitHub and can be found
on following 
[link](https://github.com/b-rodrigues/rix_paper)^[https://github.com/b-rodrigues/rix_paper].
The paper can be easily compiled by running the following command:

>>>>

---
title: "Nix for Polyglot, Reproducible Data Science Workflows"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
abstract: |
  Reproducible analysis requires more than clean, well-documented code: 
  researchers must also manage software dependencies, computational environments, and workflow execution. 
  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge, 
  but no single solution handles polyglot environments, system dependencies, 
  and workflow orchestration together.
  I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot 
  environments and declarative workflow orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define reproducible environments 
  spanning [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, and system dependencies. Building on these environments, [rixpress]{.pkg} orchestrates polyglot pipelines where each 
  computational step runs in its own hermetically sealed environment, with automatic caching and dependency tracking. 
  This approach enables researchers to develop analyses interactively while maintaining bit-for-bit reproducibility, supports 
  collaboration across heterogeneous computational environments, and ensures that analyses remain executable years into the future.

@@

However, [renv]{.pkg} has important limitations. It does not restore the version
of [R]{.proglang} itself—installing the correct R version must be done
separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
[renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg} cannot
resolve this conflict. Users must manually install system libraries, and the
process differs across operating systems. Further details are discussed in the
[renv]{.pkg} documentation^[See \url{https://rstudio.github.io/renv/articles/renv.html}.].

@@

For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, see @dolstra2004nix.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, I recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`. This shell contains all the packages specified in `default.nix` and
can be used for development, similar to activating a virtual environment in the
[Python]{.proglang} ecosystem.

@@

The `rstats-on-nix` fork allows me to circumvent these limitations. For example, it provides
daily snapshots of CRAN. Each day, the [R]{.proglang} package set is updated and
committed to a dated branch using GitHub Actions. Users can select a specific
date with:

@@

I strive to provide an available date per week: each Monday, a GitHub Action
tests popular packages on Linux and macOS, and only if all tests succeed is the
date added to the list of available dates in [rix]{.pkg}. This ensures users can
reliably install packages, and allows me to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. I
backported the fix to the `rstats-on-nix` fork, making RStudio available to users of 
[rix]{.pkg} earlier than upstream, as merging PRs in the official repository can take some
time.

@@

To illustrate these capabilities with a realistic research task, I present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang} [@bezanson2017julia], uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report. The code for this example can be found in the  
[following repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
with additional details provided in the Appendix of this manuscript.

@@

Furthermore, [rixpress]{.pkg} extends this reproducibility to entire analysis
pipelines. By leveraging Nix as a build automation engine, [rixpress]{.pkg}
allows polyglot workflows where each step runs in its own hermetically sealed
environment. This ensures deep reproducibility, efficient caching, and seamless
orchestration of polyglot analyses. Most importantly, it provides
native polyglot support without the manual coordination required by 
containerization approaches, making complex multi-language workflows 
accessible to researchers without systems administration expertise.

@@

The source code of this paper is hosted on GitHub and can be found
at following 
[link](https://github.com/b-rodrigues/rix_paper)^[https://github.com/b-rodrigues/rix_paper].
The paper can be easily compiled by running the following command:
